\thetitle

### 2020-11-18 writing/challenges.tex
revision of rational

\section{Why is it hard to evaluate blame?}
\label{sec:challenges}

@@

\item Generate a large set of programs with representative
  type errors that all three systems can detect (section~\ref{sec:mutate});

\item Simulate the rational programmer on these programs with sufficient 
  modes to (i) match the peculiarities of each  system and 
    (ii) isolate blame from other confounding factors
    (section~\ref{sec:rational});

\item Sample the experimental space  to obtain a computationally feasible
  but statistically meaningful experiment (section~\ref{sec:sample}).

>>>>

\section{Why is it hard to evaluate blame?}

@@

\item Generate a large set of programs with interesting
  type-level errors (section~\ref{sec:mutate});

\item Simulate the rational programmer on these programs with sufficient 
  modes to make meaningful comparisons between modes of the same or
    different systems
    (section~\ref{sec:rational});

\item Sample the experimental space  to obtain a computationally feasible
  but statistically sound experiment (section~\ref{sec:sample}).

### 2020-11-18 writing/rational.tex
revision of rational

Based on this idea, migratory paths offer a concise setting for
evaluating the effectiveness of a rational programmer. If we group the
scenarios from section ~\ref{sec:mutate} by mutation, then we obtain a set 
of programs that all contain the same mistake but differ in their type
annotations.  In the setting of migratory typing,\footnote{The method can be adapted to
so-called true gradual typing systems~\cite{svcb-snapl-2015} in a
straightforward manner.} each scenario differs from another in which
components are typed. Formally, a program $\system$ is a set of
components $\set{\component}$, and a scenario $\conf$ is the typed subset of
$\set{\component}$ . Hence, the scenarios of $\system$ are ordered by the
subset relation and, naturally,  form a lattice $\lattice{\system}$ with
$2^{\size{\system}}$ elements, like
the one introduced by ~\citet{tfgnvf-popl-2016} for performance
evaluations. The bottom scenario of
$\lattice{\system}$ is always $\emptyset$ and the top one is $\system$
itself. In between are all the mix-typed scenarios.

With these definitions, the actions of the rational programmer map to an
ascending chain---dubbed a \emph{trail}---in $\lattice{\system}$.  The
starting scenario $\conf_0$ of a trail---also referred to as the
\emph{root} (of the trail)---is the initial debugging scenario.  Herein we
consider only roots that are interesting debugging scenarios as described
in~\ref{sec:mutate} --- we return to this point further on.  Intuitively,
starting from such a root, the rational programmer  steps along a trail,
and at each step, uses reduces the current scenario to an ``easier'' next
scenario to debug.  The easiest scenarios are those that the type checker
rejects outright, because a type error pinpoints an impedance mismatch
between the broken code of one component and its typed interface to
others. At this point, the programmer has identified the broken component.
If the scenario type-checks, the programmers runs the program until it
raises an exception.  The rational programmer uses the blame
assignment\footnote{Blame can result either directly from a run time type
check or, in the absence of blame, from an exception that we track back to
the component that causes it with the help of the stack trace.} to decide
which component(s) to equip with types. This choice constructs scenario
$\conf_{i+1}$ from $\conf_{i}$ and implies the creation of trail.

@@

The next three subsections specialize the trail-scenario idea to the
various semantics of gradual typing. Once this machinery is in place,
we use the fourth subsection to state the precise experimental
research questions that correspond to the high-level question on the
first page and to explain our experimental process. The final
subsection extends the development with a discussion of how we can
compare the effectiveness of the rational programmer across semantics.

@@

While these definitions describe how a rational programmer might gauge whether it
pays off to heed blame assignments during debugging, they do not answer whether
blame is a critical piece of the rational programmer's process.  For instance,
typing the top of stack trace of a failed run time type check might be as useful as typing the blamed one.

To account for this situation, we devise a new mode of the natural rational
programmer that follows a migration process based on exceptions:
\begin{quote}
\it A natural exception trail is a sequence of scenarios $\conf_0,...\conf_n$ of a
program $\system$ such that for all $0 \leq i \leq n - 1$, $\conf_i \subset
\conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i = \{\random{\system, \conf_i}\}$
where $\exception{\system, \conf}$ denotes the location of the runtime
  error of $\conf$.
\end{quote}
Similar to natural blame we determine how successful the exception
migration is based on the success of natural exception trails: 
\begin{quote}
\it
Given a program $\system$, the natural exception mode is\\
(1) \emph{always useful} iff $\lattice{\system}$ contains no failing
  natural exception trails.\\
(2) \emph{sometimes useful} iff $\lattice{\system}$ contains successful
  natural exception trails.\\
(3) \emph{never useful} iff $\lattice{\system}$ contains no successful natural
exception trails.\\
\end{quote}

Putting the pieces together, we can now define that blame assignments adds value in
the context of a natural semantics if blame mode  is better than the
exception.
\begin{quote}
\it Natural blame adds value to natural for a program $\system$ over
  exceptions iff the natural blame mode is more successful than the
  natural exception mode. And, the natural blame mode is more successful
  than the natural exception mode for $\system$ if either natural blame is
  always successful for $\system$ when natural exception is sometimes or
  never successful, or natural blame is sometimes successful for $\system$
  when natural exception is never successful.
\end{quote}

@@

The transient semantics assigns blame to a series of components. In this setting, a
blame assignment has the following meaning: the witness value to the impedance
mismatch crossed the boundaries between neighbors in the sequence, and each crossing
checked a portion of the value's type shape.

This ambiguity in transient blame raises the question of how the rational programmer
should react when the language produces a blame set. Our answer is that the rational
programmer has at least two reasonable options. The first one is to select the component that is added
to the blame set first and assign types to only that one---after all the types of
this first component should be able to detect an impedance mismatch earlier in the
evaluation of a program than the later ones. The second option is to
select the component that is added to the blame set last, effectively
interpreting the blame set as an enhanced stack trace.

@@

The definition of \emph{transient exception trails} 
is analogous to that for natural and, same as for natural, we use it to  
specify when the transient-last and transient-first blame modes add
value to transient for a program.

@@

 Since gradually typed languages with erasure semantics do not come with
 blame assignment, a rational programmer can only hope that the underlying
 safety checks and their exceptions are helpful.  Thus, the erasure 
 rational programmer has a single mode; the erasure exception mode.

@@

Trails and their properties give us the tools for a rigorous examination
of blame for natural, transient and erasure Typed Racket in the setting of the 
mutants from section~\ref{sec:mutate}. In particular, our experimental
process collects data to answer three questions for each mutant:
\begin{itemize}
\item[$Q_1$] Is blame valuable for natural?

@@

@@

\begin{figure}[ht]
\center
{\begin{tabular}{l|c|c}
%% ---------------------------------------------------------------------------------------------------------------
                        & {\bf Natural}        & {\bf Transient}         \\ \hline 
%% ---------------------------------------------------------------------------------------------------------------
{\bf Blame}             &       $Q_1$          &                           \\
{\bf First blame}         &                      &     $Q_2$                \\
{\bf Last blame}   &                      &     $      Q_3$                  \\
{\bf Exceptions}         &       $Q_1$    &     $Q_2$/$Q_3$    \\
\end{tabular}}
  \caption{ Experimental questions and rational programmer modes.}
  \label{fig:experiment-outline}
\end{figure}

@@

In detail we answer $Q_1$ for an ill-typed mutant from
section~\ref{sec:mutate} by analyzing the natural blame and
natural exception trails in the scenario lattice of the mutant. A
log reports on success and failure endings.  The first step is to construct the
mutant's scenario lattice and select an interesting debugging scenario
as  the root of a trail.
Our test
bed extends these trails according to the natural-blame programmer.  If no scenarios
can be added to the trail---a trail may the top of the finite lattice---we check if
the last scenario of the trail type checks or not. If it does, this natural-blame
trail is successful, otherwise it is failing. We repeat the same process for the
same roots once more following the natural-exception programmer
. Finally, based on the collected data, we calculate the
success/failure results for the trails to determine whether natural blame is more
successful than natural exceptions  and we obtain the answer for
$Q_1$.  The process is analogous for the other questions employing the respective
modes of the rational programmer. Figure~\ref{fig:process} summarizes the
experimental process for one mode of the rational programmer and connects
it with the mutations from section~\ref{sec:mutate}.

@@

\subsection{Comparing Modes through Programmer Effort}

In addition to successful and failure information for the various trails,  
our experiment  records the number of components a rational programmer 
has to type along each successful trail ($\lvert \conf_n \setminus \conf_0
\rvert$). We use this number as the metric for the effort 
of the rational programmer to debug the root scenario of each successful trail.

Programmer effort can help shed some light to the relative effectiveness of the
three gradual typing systems. For example, consider a debugging scenario from  the
lattice of one of our mutants.  If the
scenario results in blame both under, for instance, the natural and transient semantics,
then the natural blame rational programmer and the two transient blame
rational programmers (all and firstborn) can compete to see which one
reaches an ill-typed scenario with less effort. In
general, the effort of trails of different modes that have the same
root is a point of comparison for the effectiveness of these
modes even if the modes correspond to different gradual typing systems.
Since in our experiment all roots are interesting debugging scenarios and 
those are independent of mode or semantics, effort of trails is sufficient 
to compare the relative value of one mode versus another.

>>>>

Based on this idea, migratory paths offer a concise setting for
evaluating the effectiveness of a rational programmer for dealing with the
interesting debugging scenarios  from section ~\ref{sec:mutate}.
If we group debugging scenarios by mutation, then we obtain a set 
of programs that all contain the same mistake but differ in their type
annotations.  In the setting of migratory typing,\footnote{The method can be adapted to
so-called true gradual typing systems~\cite{svcb-snapl-2015} in a
straightforward manner.} each scenario differs from another in which
components are typed. Formally, a program $\system$ is a set of
components $\set{\component}$, and a scenario $\conf$ is the typed subset of
$\set{\component}$ . Hence, the scenarios of $\system$ are ordered by the
subset relation and, naturally,  form a lattice $\lattice{\system}$ with
$2^{\size{\system}}$ elements, like
the one introduced by ~\citet{tfgnvf-popl-2016} for performance
evaluations. The bottom scenario of
$\lattice{\system}$ is always $\emptyset$ and the top one is $\system$
itself. In between are all the mix-typed scenarios.

With these definitions, the actions of the rational programmer map to an
ascending chain---dubbed a \emph{trail}---in $\lattice{\system}$.  The
starting scenario $\conf_0$ of a trail---also referred to as the
\emph{root} (of the trail)---is the initial debugging scenario.  Herein we
consider only roots that are interesting debugging scenarios as described
in~\ref{sec:mutate} --- we expand on this point further on.  Intuitively,
starting from such a root, the rational programmer  steps along a trail,
and at each step, reduces the current scenario to an ``easier'' next
scenario to debug.  The easiest scenarios are those that the type checker
rejects outright, because a type error pinpoints an impedance mismatch
between the broken code of one component and its typed interface to
others. At this point, the programmer has identified the broken component.
If the scenario type-checks, the programmers runs the program until it
raises a run time error.  The rational programmer uses the blame
assignment\footnote{Blame can result either directly from a run time type
check or, in the absence of blame, from an exception that we track back to
the component that causes it with the help of the stack trace.} to decide
which component(s) to equip with types. This choice constructs scenario
$\conf_{i+1}$ from $\conf_{i}$ and implies the creation of trail.

@@

The next three subsections specialize the trail-scenario idea to the
various semantics of gradual typing. Once this machinery is in place,
we use the fourth subsection to state the precise experimental
research questions that correspond to the high-level question on the
first page and to explain our experimental process. The final
subsection extends the development with a discussion of programmer effort.

@@

While these definitions describe how a rational programmer might gauge whether it
pays off to heed blame assignments during debugging, they do not answer whether
blame is a critical piece of the rational programmer's process.  For instance,
typing the top of stack trace of a failed run time type check, dubbed the
location of the exception, might be as useful as typing the blamed one.

To account for this situation, we devise a new mode of the natural rational
programmer that follows a migration process based on exceptions:
\begin{quote}
\it A natural exception trail is a sequence of scenarios $\conf_0,...\conf_n$ of a
program $\system$ such that for all $0 \leq i \leq n - 1$, $\conf_i \subset
\conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i = \{\random{\system, \conf_i}\}$
where $\exception{\system, \conf}$ denotes the location of the exception of $\conf$.
\end{quote}
Similar to natural blame we determine how successful the exception
migration is based on the success of natural exception trails: 
\begin{quote}
\it
Given a program $\system$, the natural exception mode is\\
(1) \emph{always useful} iff $\lattice{\system}$ contains no failing
  natural exception trails.\\
(2) \emph{sometimes useful} iff $\lattice{\system}$ contains successful
  natural exception trails.\\
(3) \emph{never useful} iff $\lattice{\system}$ contains no successful natural
exception trails.\\
\end{quote}

Putting the pieces together, we can now define when blame 
is more useful than mere exceptions in the context of a natural semantics.
\begin{quote}
\it Natural blame  is more useful
  than natural exceptions for $\system$ if either natural blame is
  always useful for $\system$ when natural exception is sometimes or
  never useful, or natural blame is sometimes useful for $\system$
  when natural exception is never useful.
\end{quote}

@@

The transient semantics assigns blame to a set of components. In this setting, a
blame assignment has the following meaning: the witness value to the impedance
mismatch may have crossed the boundaries between elements in the set, and each crossing
checked only a portion of the value's type shape.

This ambiguity in transient blame raises the question of how the rational programmer
should react when the language produces a blame set. Our answer is that the rational
programmer has at least two reasonable options. The first one is to select the component that is added
to the blame set first and assign types to only that one---after all, if
fully checked, the types of
this first component should be able to detect an impedance mismatch earlier in the
evaluation of a program than the later ones. The second option is to
select the component that is added to the blame set last, effectively
interpreting the blame set as a boundary-aware stack.

@@

The definition of \emph{transient exception trails} 
is analogous to that for natural and, same as for natural, we use it to  
specify when the transient-last and transient-first blame modes are more
useful than exceptions in the context of a natural semantics.

@@

 Since gradually typed languages with erasure semantics do not come with
 blame assignment, a rational programmer can only hope that the underlying
 safety checks and their exceptions are helpful.  Thus, the erasure 
 rational programmer has a single mode, the erasure exception mode, and
 its definition and the accompanying notion of usefulness follow that for
 natural.

@@

Trails and their properties give us the tools for a rigorous examination
of blame for natural, transient and erasure Typed Racket in the setting of the 
mutants from section~\ref{sec:mutate}. In line with the discussion so far, 
our experimental process collects data to answer three initial questions for each mutant:
\begin{itemize}
\item[$Q_1$] Is blame valuable for natural?

@@

Furthermore, our experiment can compare the relative value of blame
between any two of the three semantics. 
\begin{itemize}
\item[$Q_*$] Is blame for X more valuable than blame for Y where X, Y is natural,
  erasure or natural?
\end{itemize}

@@

\begin{figure}[ht]
\center
{\begin{tabular}{l|c|c|c}
%% ---------------------------------------------------------------------------------------------------------------
                        & {\bf Natural}  & {\bf Transient} &  {\bf Erasure} \\ \hline 
%% ---------------------------------------------------------------------------------------------------------------
{\bf Blame}             &  $Q_1/Q_*$    &                  &                \\
{\bf First blame}       &               &     $Q_2/Q_*$    &                 \\
{\bf Last blame}        &               &     $Q_3/Q_*$    &                 \\
{\bf Exceptions}        &      $Q_1$    &     $Q_2/Q_3$    &      $Q_*$      \\
\end{tabular}}
  \caption{ Experimental questions and rational programmer modes.}
  \label{fig:experiment-outline}
\end{figure}

@@

In detail we answer $Q_1$ for an ill-typed mutant from
section~\ref{sec:mutate} by analyzing the natural blame and
natural exception trails in the scenario lattice of the mutant. A
log reports on success and failure endings.  The first step is to construct the
mutant's scenario lattice and select one of the interesting debugging
scenarios from section~\ref{sec:mutate}.
Our test
bed extends these trails according to the natural-blame programmer.  If no scenarios
can be added to the trail, we check if
the last scenario of the trail type checks or not. If it does, this natural-blame
trail is successful, otherwise it is failing. We repeat the same process for the
same roots once more following the natural-exception programmer. 
Next, based on the collected data, we calculate the
success/failure results for the trails to determine whether blame
is more useful than exceptions for this mutant. Finally, we
obtain a positive answer for
$Q_1$ when the percentage of mutants where blame is more useful
than exceptions is greater than the percentage of mutants where exceptions
are more useful
than blame.  The process is analogous for $Q_2$ and $Q_3$  employing the respective
modes of the rational programmer. For $Q_*$, we also proceed in a similar
manner except that we use modes 
of the rational programmer from different semantics. For instance to
decide whether blame for natural is more valuable than blame for
transient, we compare the usefulness of
the natural blame mode with that of the transient first and transient last
modes for all our mutants. 
Figure~\ref{fig:process} summarizes the
experimental process for one mode of the rational programmer and connects
it with the mutations from section~\ref{sec:mutate}.

@@

\subsection{Programmer Effort}

In addition to successful and failure information for the various trails,  
our experiment  records the number of components a rational programmer 
has to type along each trail ($\lvert \conf_n \setminus \conf_0
\rvert$). We use this number as the metric for the effort 
of the rational programmer to debug an interesting scenario.

A comparison of the effort of different modes of the rational programmer
can help shed some further light to the effectiveness of the
three gradual typing systems. For example, consider one of 
our mutants.  If, for instance, both natural blame and transient first
blame are equally useful for the mutant, the two nodes of the 
rational programmer can compete to see which one
debugs the interesting scenarios of the mutant with less effort. In
general, if the effort distribution of a mode of the rational programmer for a
mutant has a shorter tail and more
volume around smaller values compared to the effort distribution of another
mode, then the first mode is the more effective of the two.

Finally, effort allows us to validate that the observed effectiveness 
of the rational programmer is not an artifact of pure chance.
To test the null hypothesis for the effectiveness of the rational
programmer, we can simply compare the effort distribution of a mode of the
rational programmer with a rational programmer that ignores blame and
exceptions and instead selects which component to type next randomly. 
Of course, since each mutant has a finite number of components, 
such a random mode is always useful for all mutants in the context of all
three semantics. At the same time though,
the natural, transient and erasure random mode exhibit a normal effort distribution for each mutant
that should be distinct from the effort distribution of all other modes of
the rational programmer if the effectiveness of the latter is not
coincidental.
